A Solution to Static vs. Dynamic Linking

### 2024-04-10 mss.tex
Post proof.

\newcommand{\authorname}{Ted Blackman \patp{rovnys-ricfer} \\ Philip C. Monk \patp{wicdev-wisryt}}
\newcommand{\authorpatp}{}
\newcommand{\affiliation}{Urbit Foundation \\ Tlon Corporation}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \patp{rovnys-ricfer}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Static vs. Dynamic Linking}
\fancyfoot[LE,RO]{\thepage}

@@

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{1}

@@

It’s now such an enormous problem there are multiple billion-dollar startups for technologies for dealing with this complexity by adding further complexity. Docker, Kubernetes, various Amazon atrocities for dealing with Docker and Kubernetes and their competitors, Flatpak, Conda, AppImages, MacPorts, \texttt{brew}, RPM, NixOS, \texttt{dpkg}/\texttt{apt}, VirtualBox, \texttt{pacman}, Yum, SnapCraft…. As the number of packages grows, this breaks down, and even the OS maintainers are giving up and turning to flatpak, AppImages and Snap files. 
These are extremely complicated and incredibly wasteful \emph{(of memory)} ways of literally packaging up a bunch of needed shared libraries with your application and presenting it to you as a crappy simulacrum of a statically compiled binary.  \citep{Locklin2022}
\end{quote}

\noindent
\citeauthor{Locklin2022}'s picturesque exposition highlights the “dependency hell” or “DLL hell” that mires modern software development (cf. \citet{Grimes2003}).  From Urbit's perspective as a solid-state computer, another problem with dynamic linking is that it is not deterministic.  Dynamic linking was something of a pact with the devil, permitting efficient memory usage at the price of legibility.

@@

Determinism has long been a desired characteristic of any given build system.  The source code should be a function of build environment and build instructions in a straightforward way.  Even accessing the linked libraries in a different order can alter the resulting binary, however, meaning that true reproducibility is elusive.  Declarative package managers like Nix \citep{Nix}~and Guix \citep{Guix}~use a functional package management approach to achieve reproducibility, marking packages using cryptographic hashes to track dependencies uniquely and repeatably.

@@

As another example, the Hoon source file with text \lstinline[style=inlinecode]{'[3 0x4]'} compiles to
\begin{lstlisting}[style=listingblock]
[[%cell [%atom %ud ~] [%atom %ux ~]] [3 0x4]]
\end{lstlisting}
\noindent
This is a vase of a \lstinline[style=inlinecode]{%cell} (pair) of unsigned decimal number (\lstinline[style=inlinecode]{%ud}) and unsigned hexadecimal number (\lstinline[style=inlinecode]{%ux}).  If \lstinline[style=inlinecode]{/foo/hoon} is \lstinline[style=inlinecode]{'3'}, \lstinline[style=inlinecode]{/bar/hoon} is \lstinline[style=inlinecode]{'0x4'}, then importing both of those files changes the build subject to (a vase of) \lstinline[style=inlinecode]{[foo=3 bar=0x4 <stdlib>]}.

@@

The former Ford cache was per-desk,\footnote{Clay desks in Urbit are like Git branches, describing particular filesystem continuities.} keyed on the name of the build (e.g. a file at a certain path).  It was impossible to share such a cache between desks because the name may refer to different things on different desks or at different revisions.

@@

Reclaiming space from this cache becomes important.  Since this is a “true” cache, it's never incorrect to keep data in it.  One could adopt a heuristic such as to clear hourly or so you could use a heuristic such as "clear the whole thing every hour" or a least-recently-used (LRU) policy.  However, \lstinline[style=inlinecode]{++ford} has a long history of trying to use such heuristics and still using exorbitant amounts of memory.  The primary innovation of the former \lstinline[style=inlinecode]{++ford} cache was that its size is deterministic, and it stored no builds unless they could be generated from the head of its desk (\lstinline[style=inlinecode]{%home} then \lstinline[style=inlinecode]{%base}).

>>>>

\newcommand{\authorname}{Ted Blackman \patp{rovnys-ricfer} \\ Urbit Foundation}
\newcommand{\authorpatp}{}
\newcommand{\affiliation}{Philip C. Monk \patp{wicdev-wisryt} \\ Tlon Corporation}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  75–82. \\
              ~ \\
              Address author correspondence to \patp{rovnys-ricfer}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Static vs. Dynamic Linking}
\fancyfoot[LE,RO]{\thepage}

@@

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{75}

@@

\sloppy
It's now such an enormous problem there are multiple billion-dollar startups for technologies for dealing with this complexity by adding further complexity. Docker, Kubernetes, various Amazon atrocities for dealing with Docker and Kubernetes and their competitors, Flatpak, Conda, AppImages, MacPorts, \texttt{brew}, {\sc rpm}, NixOS, \texttt{dpkg}/\texttt{apt}, VirtualBox, \texttt{pacman}, Yum, SnapCraft…. As the number of packages grows, this breaks down, and even the OS maintainers are giving up and turning to flatpak, AppImages and Snap files.

\sloppy
These are extremely complicated and incredibly wasteful \emph{(of memory)} ways of literally packaging up a bunch of needed shared libraries with your application and presenting it to you as a crappy simulacrum of a statically compiled binary.  \citep{Locklin2022}
\end{quote}

\noindent
\citeauthor{Locklin2022}'s picturesque exposition highlights the “dependency hell” or “{\sc dll} hell” that mires modern software development (cf. \citet{Grimes2003}).  From Urbit's perspective as a solid-state computer, another problem with dynamic linking is that it is not deterministic.  Dynamic linking was something of a pact with the devil, permitting efficient memory usage at the price of legibility.

@@

\sloppy
Determinism has long been a desired characteristic of any given build system.  The source code should be a function of build environment and build instructions in a straightforward way.  Even accessing the linked libraries in a different order can alter the resulting binary, however, meaning that true reproducibility is elusive.  Declarative package managers like Nix \citep{Nix}~and Guix \citep{Guix}~use a functional package management approach to achieve reproducibility, marking packages using cryptographic hashes to track dependencies uniquely and repeatably.

@@

As another example, the Hoon source file with text \lstinline[style=inlinecode]{'[3 0x4]'} compiles to
\begin{lstlisting}[style=listingblock]
[[%cell [%atom %ud ~] [%atom %ux ~]] [3 0x4]]
\end{lstlisting}
\noindent
This is a vase of a \lstinline[style=inlinecode]{%cell} (pair) of unsigned decimal number (\lstinline[style=inlinecode]{%ud}) and unsigned hexadecimal number (\lstinline[style=inlinecode]{%ux}).  If \lstinline[style=inlinecode]{/foo/hoon} is \lstinline[style=inlinecode]{'3'}, \lstinline[style=inlinecode]{/bar/hoon} is \lstinline[style=inlinecode]{'0x4'}, then importing both of those files changes the build subject to \lstinline[style=inlinecode]{[foo=3 bar=0x4 <stdlib>]} (or really, a vase of that value).

@@

\sloppy
The former Ford cache was per-desk,\footnote{Clay desks in Urbit are like Git branches, describing particular filesystem continuities.} keyed on the name of the build (e.g. a file at a certain path).  It was impossible to share such a cache between desks because the name may refer to different things on different desks or at different revisions.

@@

Reclaiming space from this cache becomes important.  Since this is a “true” cache, it's never incorrect to keep data in it.  One could adopt a heuristic such as to clear hourly or so you could use a heuristic such as "clear the whole thing every hour" or a least-recently-used ({\sc lru}) policy.  However, \lstinline[style=inlinecode]{++ford} has a long history of trying to use such heuristics and still using exorbitant amounts of memory.  The primary innovation of the former \lstinline[style=inlinecode]{++ford} cache was that its size is deterministic, and it stored no builds unless they could be generated from the head of its desk (\lstinline[style=inlinecode]{%home} then \lstinline[style=inlinecode]{%base}).
